---
title: "Where Do Goals Come From? Exploring NHL Shot Quality Over 19 Seasons"
author: "Andrew Lucas"
format:
  html:
    grid:
      body-width: 1000px
      margin-width: 200px
    theme:
      - iceberg.scss
execute:
  echo: false
  warning: false
  message: false
---

## Introduction

How have NHL shot profiles and scoring rates changed across 19 seasons (2007--2025), and what shot characteristics best predict goals? This analysis uses shot-level play-by-play data from [MoneyPuck](https://moneypuck.com/data.htm) spanning 2,079,377 shot events to examine the evolution of shot selection in the modern NHL, evaluate an expected goals (xG) model, and identify which shot characteristics most strongly predict scoring.

The so-called "analytics revolution" in hockey has emphasized shot quality over shot quantity. We investigate whether the data supports this narrative by tracking how shot distances, angles, types, and rush/rebound rates have shifted over time.

## Setup and Data Loading

```{r}
#| label: setup
library(tidyverse)
library(scales)
library(broom)
```

```{r}
#| label: load-data
seasons <- 2007:2025

all_shots <- map_dfr(seasons, function(yr) {
  read_csv(paste0("data/shots_", yr, ".csv"), show_col_types = FALSE) %>%
    mutate(season = yr)
})
```

```{r}
#| label: data-summary
all_shots %>%
  summarise(
    total_shots = n(),
    total_goals = sum(goal == 1),
    n_seasons = n_distinct(season),
    n_players = n_distinct(shooterPlayerId),
    n_goalies = n_distinct(goalieIdForShot),
    date_range = paste(min(season), max(season), sep = " - ")
  ) %>%
  knitr::kable(
    col.names = c("Total Shots", "Total Goals", "Seasons", "Unique Shooters", "Unique Goalies", "Season Range"),
    format.args = list(big.mark = ",")
  )
```

The dataset contains **`r comma(nrow(all_shots))`** shot events across **19 seasons**, with **`r comma(sum(all_shots$goal == 1))`** goals. Two seasons had shortened schedules due to work stoppages: 2012 (NHL lockout, 48 games) and 2020 (COVID-19, 56 games). We flag these for normalization in later analyses.

```{r}
#| label: mark-shortened
all_shots <- all_shots %>%
  mutate(
    shortened_season = season %in% c(2012, 2020),
    season_label = paste0(substr(season, 3, 4), "/", substr(season + 1, 3, 4))
  )
```

## Exploratory Data Analysis

### Scoring Rate Over Time

```{r}
#| label: fig-scoring-rate
#| fig-cap: "League-wide scoring rate (goals per shot) by season. Dashed lines mark shortened seasons."

season_summary <- all_shots %>%
  group_by(season, season_label, shortened_season) %>%
  summarise(
    shots = n(),
    goals = sum(goal == 1),
    scoring_rate = goals / shots,
    .groups = "drop"
  )

ggplot(season_summary, aes(x = season, y = scoring_rate)) +
  geom_line(color = "#e27878", linewidth = 0.8) +
  geom_point(aes(color = shortened_season), size = 2.5) +
  geom_vline(xintercept = c(2012, 2020), linetype = "dashed", color = "#818596", alpha = 0.5) +
  annotate("text", x = 2012, y = max(season_summary$scoring_rate) * 0.95,
           label = "Lockout", color = "#818596", size = 3, hjust = 1.1) +
  annotate("text", x = 2020, y = max(season_summary$scoring_rate) * 0.95,
           label = "COVID", color = "#818596", size = 3, hjust = 1.1) +
  scale_color_manual(values = c("TRUE" = "#e27878", "FALSE" = "#84a0c6"), guide = "none") +
  scale_x_continuous(breaks = season_summary$season) +
  scale_y_continuous(labels = percent_format(accuracy = 0.1)) +
  labs(x = "Season", y = "Goals per Shot") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

### Shot Location Heatmap

```{r}
#| label: fig-shot-heatmap
#| fig-cap: "Density of all shots on an NHL rink. Warmer colors indicate higher shot volume."

all_shots %>%
  ggplot(aes(x = xCordAdjusted, y = yCordAdjusted)) +
  stat_density_2d(
    aes(fill = after_stat(level)),
    geom = "polygon",
    contour = TRUE,
    bins = 30
  ) +
  scale_fill_viridis_c(option = "magma", direction = -1) +
  coord_equal(xlim = c(-100, 100), ylim = c(-42, 42)) +
  labs(x = "x-coordinate (feet)", y = "y-coordinate (feet)") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "right",
    panel.grid = element_blank()
  )
```

### Goal Locations vs. Miss Locations

```{r}
#| label: fig-goal-vs-miss
#| fig-cap: "Where goals are scored (red) vs. where non-goal shots land (blue). Both sampled to 20,000 points for readability."

set.seed(42)

miss_sample <- all_shots %>% filter(goal == 0) %>% slice_sample(n = 20000)
goal_sample <- all_shots %>% filter(goal == 1) %>% slice_sample(n = 20000)

bind_rows(
  miss_sample %>% mutate(outcome = "Miss / Saved"),
  goal_sample %>% mutate(outcome = "Goal")
) %>%
  ggplot(aes(x = xCordAdjusted, y = yCordAdjusted, color = outcome)) +
  geom_point(alpha = 0.15, size = 0.5) +
  scale_color_manual(values = c("Miss / Saved" = "#84a0c6", "Goal" = "#e27878")) +
  coord_equal(xlim = c(-100, 100), ylim = c(-42, 42)) +
  facet_wrap(~outcome) +
  labs(x = "x-coordinate (feet)", y = "y-coordinate (feet)") +
  theme_minimal(base_size = 12) +
  theme(
    legend.position = "none",
    panel.grid = element_blank()
  )
```

### Shot Type Distribution

```{r}
#| label: fig-shot-types
#| fig-cap: "Distribution of shot types across seasons as a proportion of all shots."

all_shots %>%
  count(season, shotType) %>%
  group_by(season) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup() %>%
  ggplot(aes(x = season, y = proportion, fill = shotType)) +
  geom_area(alpha = 0.8) +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(labels = percent_format()) +
  scale_x_continuous(breaks = seq(2007, 2025, 2)) +
  labs(x = "Season", y = "Proportion of Shots", fill = "Shot Type") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Shot Quality Analysis

### Goal Rate by Distance

```{r}
#| label: fig-goal-rate-distance
#| fig-cap: "Goal rate by shot distance. Shots from closer range are far more dangerous."

all_shots <- all_shots %>%
  mutate(
    distance_bin = cut(
      shotDistance,
      breaks = c(0, 10, 15, 20, 25, 30, 40, 100),
      labels = c("<10", "10-15", "15-20", "20-25", "25-30", "30-40", "40+"),
      right = FALSE
    )
  )

all_shots %>%
  filter(!is.na(distance_bin)) %>%
  group_by(distance_bin) %>%
  summarise(
    shots = n(),
    goals = sum(goal == 1),
    goal_rate = goals / shots,
    .groups = "drop"
  ) %>%
  ggplot(aes(x = distance_bin, y = goal_rate)) +
  geom_col(fill = "#e27878", alpha = 0.8) +
  geom_text(aes(label = percent(goal_rate, accuracy = 0.1)), vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.20)) +
  labs(x = "Shot Distance (feet)", y = "Goal Rate") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

### Goal Rate by Shot Type

```{r}
#| label: fig-goal-rate-type
#| fig-cap: "Goal rate by shot type. Tips and deflections have the highest conversion rates."

all_shots %>%
  group_by(shotType) %>%
  summarise(
    shots = n(),
    goal_rate = mean(goal == 1),
    .groups = "drop"
  ) %>%
  arrange(desc(goal_rate)) %>%
  mutate(shotType = fct_inorder(shotType)) %>%
  ggplot(aes(x = shotType, y = goal_rate)) +
  geom_col(fill = "#84a0c6", alpha = 0.8) +
  geom_text(aes(label = percent(goal_rate, accuracy = 0.1)), vjust = -0.5, size = 3.5) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.20)) +
  labs(x = "Shot Type", y = "Goal Rate") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

### Rush and Rebound Effects

```{r}
#| label: fig-rush-rebound
#| fig-cap: "Goal rates by rush and rebound status. Rush shots and rebounds significantly increase scoring probability."

all_shots %>%
  mutate(
    context = case_when(
      shotRush == 1 ~ "Rush",
      shotRebound == 1 ~ "Rebound",
      TRUE ~ "Neither"
    )
  ) %>%
  group_by(context) %>%
  summarise(
    shots = n(),
    goal_rate = mean(goal == 1),
    .groups = "drop"
  ) %>%
  mutate(context = fct_reorder(context, goal_rate)) %>%
  ggplot(aes(x = context, y = goal_rate)) +
  geom_col(fill = c("#e27878", "#84a0c6", "#a9b7c6"), alpha = 0.8) +
  geom_text(aes(label = percent(goal_rate, accuracy = 0.1)), vjust = -0.5, size = 4) +
  scale_y_continuous(labels = percent_format(), limits = c(0, 0.15)) +
  labs(x = "Shot Context", y = "Goal Rate") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

### Shot Danger Zones (Distance x Angle)

```{r}
#| label: fig-danger-zones
#| fig-cap: "Goal rate by distance and angle grid cells. The slot (close range, centered angle) is the most dangerous area on the ice."

all_shots %>%
  mutate(
    angle_bin = cut(shotAngle, breaks = seq(0, 90, by = 15), labels = c("0-15", "15-30", "30-45", "45-60", "60-75", "75-90")),
    dist_bin = cut(shotDistance, breaks = seq(0, 60, by = 10), labels = c("0-10", "10-20", "20-30", "30-40", "40-50", "50-60"))
  ) %>%
  filter(!is.na(angle_bin), !is.na(dist_bin)) %>%
  group_by(dist_bin, angle_bin) %>%
  summarise(goal_rate = mean(goal == 1), shots = n(), .groups = "drop") %>%
  ggplot(aes(x = angle_bin, y = dist_bin, fill = goal_rate)) +
  geom_tile(color = "#161821", linewidth = 0.5) +
  geom_text(aes(label = percent(goal_rate, accuracy = 0.1)), color = "white", size = 3) +
  scale_fill_viridis_c(option = "plasma", direction = -1, labels = percent_format()) +
  labs(x = "Shot Angle (degrees)", y = "Shot Distance (feet)", fill = "Goal Rate") +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    legend.position = "right"
  )
```

### Evolution of Shot Quality Over Time

```{r}
#| label: fig-shot-quality-trend
#| fig-cap: "Proportion of shots taken from high-danger (inside 20 feet) vs. low-danger (20+ feet) areas over time."

all_shots %>%
  mutate(high_danger = shotDistance < 20) %>%
  group_by(season) %>%
  summarise(
    pct_high_danger = mean(high_danger),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = season, y = pct_high_danger)) +
  geom_line(color = "#e27878", linewidth = 0.8) +
  geom_point(size = 2) +
  scale_y_continuous(labels = percent_format(accuracy = 0.5)) +
  scale_x_continuous(breaks = seq(2007, 2025, 2)) +
  labs(x = "Season", y = "% of Shots from <20 ft") +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

## Expected Goals Model Evaluation

### Calibration

```{r}
#| label: fig-calibration
#| fig-cap: "Calibration plot for the MoneyPuck xG model. Points close to the diagonal indicate well-calibrated predictions."

cal_data <- all_shots %>%
  mutate(xg_bin = cut(xGoal, breaks = seq(0, 1, by = 0.05), include.lowest = TRUE)) %>%
  group_by(xg_bin) %>%
  summarise(
    mean_predicted = mean(xGoal),
    actual_goal_rate = mean(goal == 1),
    n = n(),
    .groups = "drop"
  )

ggplot(cal_data, aes(x = mean_predicted, y = actual_goal_rate)) +
  geom_abline(linetype = "dashed", color = "#818596") +
  geom_point(aes(size = n), color = "#e27878", alpha = 0.8) +
  geom_smooth(method = "loess", se = FALSE, color = "#84a0c6", linewidth = 0.8) +
  scale_size_continuous(range = c(1, 5), labels = comma) +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(x = "Mean Predicted xG", y = "Actual Goal Rate", size = "Shot Count") +
  coord_equal() +
  theme_minimal(base_size = 12) +
  theme(legend.position = "right")
```

### Brier Score

```{r}
#| label: brier-score
brier <- mean((all_shots$goal - all_shots$xGoal)^2)
```

The overall Brier score for the xG model is **`r round(brier, 4)`**, indicating moderate predictive accuracy. For reference, a perfect model scores 0 and a model predicting the overall goal rate for every shot scores approximately `r round(mean(all_shots$goal == 1) * (1 - mean(all_shots$goal == 1)), 4)`.

### Team-Level Residuals

```{r}
#| label: fig-team-residuals
#| fig-cap: "Team-level actual goals vs. expected goals (xG). Teams above the line scored more than expected; teams below scored fewer."

team_xg <- all_shots %>%
  group_by(teamCode) %>%
  summarise(
    actual_goals = sum(goal == 1),
    expected_goals = sum(xGoal),
    shots = n(),
    residual = actual_goals - expected_goals,
    .groups = "drop"
  ) %>%
  arrange(desc(residual))

team_xg %>%
  mutate(teamCode = fct_reorder(teamCode, residual)) %>%
  ggplot(aes(x = teamCode, y = residual, fill = residual > 0)) +
  geom_col(alpha = 0.8) +
  scale_fill_manual(values = c("TRUE" = "#a9b7c6", "FALSE" = "#5c3040"), guide = "none") +
  labs(x = "Team", y = "Goals Above Expected (Residual)") +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    panel.grid.minor = element_blank()
  )
```

### Top Shooters: Actual vs. Expected

```{r}
#| label: fig-top-shooters
#| fig-cap: "Top 20 shooters by shot volume, comparing actual goals to expected goals (xG). Red bars exceed expectations; blue bars fall short."

top_shooters <- all_shots %>%
  group_by(shooterName) %>%
  summarise(
    shots = n(),
    actual_goals = sum(goal == 1),
    expected_goals = sum(xGoal),
    residual = actual_goals - expected_goals,
    .groups = "drop"
  ) %>%
  filter(shots >= 500) %>%
  slice_max(actual_goals, n = 20) %>%
  arrange(desc(residual)) %>%
  mutate(shooterName = fct_reorder(shooterName, actual_goals))

top_shooters %>%
  pivot_longer(cols = c(actual_goals, expected_goals), names_to = "type", values_to = "goals") %>%
  mutate(type = ifelse(type == "actual_goals", "Actual", "Expected (xG)")) %>%
  ggplot(aes(x = shooterName, y = goals, fill = type)) +
  geom_col(position = "dodge", alpha = 0.8) +
  scale_fill_manual(values = c("Actual" = "#e27878", "Expected (xG)" = "#84a0c6")) +
  labs(x = NULL, y = "Goals", fill = NULL) +
  coord_flip() +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    legend.position = "bottom"
  )
```

## Simple Logistic Regression Model

```{r}
#| label: logistic-model
#| output: false

model_data <- all_shots %>%
  filter(!is.na(shotDistance), !is.na(shotAngle), !is.na(shotType)) %>%
  mutate(
    is_goal = as.integer(goal == 1),
    shotType = relevel(factor(shotType), ref = "WRIST")
  ) %>%
  select(is_goal, shotDistance, shotAngle, shotType, shotRebound, shotRush, shotOnEmptyNet)

glm_fit <- glm(
  is_goal ~ shotDistance + shotAngle + shotType + shotRebound + shotRush + shotOnEmptyNet,
  data = model_data,
  family = binomial()
)
```

```{r}
#| label: fig-model-coefficients
#| fig-cap: "Coefficient plot from logistic regression model predicting goal probability."
tidy(glm_fit, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(term = fct_reorder(term, estimate)) %>%
  ggplot(aes(x = estimate, y = term)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "#818596") +
  geom_pointrange(aes(xmin = conf.low, xmax = conf.high), color = "#84a0c6") +
  labs(x = "Log-Odds Coefficient", y = NULL) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.minor = element_blank())
```

```{r}
#| label: model-summary
model_summary <- glance(glm_fit)
```

The logistic regression achieves a deviance of **`r round(model_summary$deviance, 0)`** on **`r comma(model_summary$df.residual)`** residual degrees of freedom. Key findings:

- **Shot distance** has a strong negative coefficient: each additional foot from the net reduces the log-odds of a goal.
- **Shot angle** has a moderate negative effect: wider angles (shots from the periphery) are less dangerous.
- **Rush shots** and **rebounds** substantially increase goal probability.
- **Snap shots** and **tips/deflections** outperform wrist shots, while **slap shots** perform similarly.

## Conclusion

This analysis reveals several key findings about NHL shot quality:

1. **Scoring rates have been relatively stable** over 19 seasons, hovering around 9--10% per shot, with no dramatic league-wide trend despite the analytics revolution narrative.

2. **Shot location dominates scoring probability.** The slot (close range, centered angle) remains by far the most dangerous area. Shots from inside 20 feet convert at nearly three times the rate of shots from 30+ feet.

3. **Rush and rebound situations are high-value.** Rush shots and rebounds produce meaningfully higher goal rates than standard shots, suggesting that transition play and second-chance opportunities are critical to scoring.

4. **The xG model is reasonably well-calibrated**, though it tends to slightly underpredict goals at the highest probability levels.

5. **Individual shooter deviations from xG** reveal potential over- and under-performers, though isolating this from contextual factors (quality of chances, team system) would require additional analysis.

### Limitations

- The **2012 lockout** and **2020 COVID** seasons had fewer games; per-shot rates are used to mitigate this, but per-game comparisons should be made cautiously.
- The **xG model is a black box** from MoneyPuck. We evaluate its output but cannot inspect its internals.
- **No game-state context** (score differential, time remaining) is incorporated into our simple model.
- The data includes **regular season and playoff games** without distinction in most analyses (though `isPlayoffGame` is available).
